Detecting distributional differences between temporal granularities for exploratory time series analysis

### 2021-06-02 paper/main.Rmd
fix graph labels

Exploratory data analysis, as coined by John W. Tukey [@Tukey1977-jx] involves many iterations of finding structures and patterns that allow the data to be informative. With temporal data available at finer scales, exploring time series data can become overwhelming with so many possible cyclic temporal granularities [@Bettini1998-ed, @Gupta2020-vo], which are temporal deconstructions that represent cyclic repetitions in time, e.g. hour-of-day, day-of-month, or any public holidays. These granularities form ordered (for eg. day-of-week where Tue is always followed by Wed, which again is followed by Thu and so on) or unordered categorical variables. Therefore, exploring univariate time series data amounts to exploring the distribution of the measured variable across different categories of the cyclic granularities. \noindent Take the example of the  electricity smart meter data used in @wang2020calendar for four households in Melbourne, Australia. Figure \ref{fig:onegran} shows the distribution of energy usage of one household (id 2) across cyclic granularity a) hour-of-day and b) month-of-year. Figure \ref{fig:onegran}(a) shows that energy consumption is higher during the morning hours (5-8) when members in the household wake up and then again in the evening hours (17-20) possibly when members get back from work with maximum variation (large inter-quartile range) in behavior in the afternoon hours (12-16). Figure \ref{fig:onegran}(b) shows the distribution of energy consumption across months January to June. The median and quartile deviation of energy usage in Jan and Feb are generally on a much higher side, possibly due to the usage of air conditioners (Jan, Feb are peak summer in Australia), however for other months (Mar-Jun, autumn and winter), the smaller median and quartile deviation indicate a more consistent behavior. It might also imply that this household do not use as much heater as compared to air conditioner. A lot of households in Victoria use gas heating and hence the usage of heaters might not be reflected here. Potentially many such displays could be drawn across day-of-week, day-of-month, weekday/weekend, or any other chosen cyclic granularities of interest. However, all of them would not be interesting to discern important patterns in the energy usage. Only those displays, which have "significant" distributional differences between categories of the cyclic granularity would be informative.

@@

```{r onegran, fig.cap="Boxplots showing the distribution of one household across one cyclic granularity at a time - (a) hour-of-day and (b) and month-of-year. The daily and annual periodic behavior of energy is apparent in (a) and (b) respectively, with daily peaks occuring in morning and evening hours when members in the house are present and active, more volatility (usage of air conditioner) in summer months (Jan, Feb) due to air conditioners and more consistent behavior in winter and autumn months (Mar, Apr, May, June)."}
```

<!-- comparing one and two grans for one households -->
\noindent Exploring the distribution of the measured variable across two cyclic granularities tend to provide more detailed information on its structure. For example, Figure \ref{fig:id2-new}(a) slice down further by showing the usage distribution across hour-of-day conditional on month-of-year across two households (id 2 and 4). It shows the hourly usage over a day do not remain same across months. Unlike other months, the 75th and 90th percentile for all hours of the day in January are high, pretty close and are not characterized by a morning and evening peak. The household in Figure \ref{fig:id2-new}(b) has 90th percentile consumption higher in summer months relative to autumn or winter, but the 75th and 90th percentile are far apart in all months, implying that the second household resorts to air conditioning much less regularly than the first one. The differences seem to be more prominent across month-of-year (facets) than hour-of-day (x-axis) for this household, whereas they are prominent for both cyclic granularities for the first household. It could be immensely useful to make the transition from all potential displays to the ones that are informative across atleast one cyclic granularity.

@@

<!-- \ref{fig:id2-new} shows the distribution of energy consumption across hour-of-day conditional on month-of-year. \ref{fig:id2-new}(a) shows that energy consumption is higher during the morning hours when members in the household wake up and then again higher in the evening hours possibly when members are back from work with them showing maximum variable behavior in the afternoon hours. \ref{fig:id2-new}(b) shows the months Jan and Feb not only have a high median consumption, but also more variability, possibly due to the usage of air conditioning. The other months shows more consistent behavior with the energy consumption rising in May and June. \ref{fig:id2-new} adds to this information, by showing that hours in Jan do not necessarily follow the same pattern across the day (due to Holidays, people visiting). Moreover, the volatility in consumption are almost equal and high for all hours in Jan, whereas variability is most in the evening hours for other months. The distribution of consumption across different hours of the day look similar for Apr, May and Jun, characterized by a distinct morning, afternoon and evening peaks and higher variability in evening hours. -->

@@

```{r id2-new, fig.cap = "Distribution of energy consumption displayed through area quantile plots across two cyclic granularities month-of-year and hour-of-day and two households. The black line is the median, whereas the orange band covers the 25th to 75th percentile and the green band covers the 10th to 90th percentile. Difference between the 90th and 75th quantiles is less for (Jan, Feb) for the first household (a), suggesting that it is a more frequent user of air conditioner than the second household (b). Energy consumption for in (a) changes across both granularities, whereas for (b) daily pattern stays same irrespective of the months."}
```

@@

```{r tab-demography, eval = FALSE}
demography <- tibble(id = c(1, 2, 3, 4, 5,6, 7, 8), 
       profession = c("academia", "mixed", "industry", "industry", "mixed", "mixed", "academia", "industry"),
       total_members = c(2, 3, 5, 5, 2, 2, 2, 6),
       kids = c("no", "no", "yes", "yes", "no", "no", "no", "yes"),
       old_parents = c("no", "yes", "yes", "yes", "no", "no", "no", "yes"),
       PhD_student = c("no", "yes", "no", "no", "yes", "yes", "yes", "no"))

demography <- demography %>% kable(format  = "markdown", caption = "Demographics of the eight households chosen for study")

demography %>% add_footnote(label = "***: 99% significant, **: 95% and *: 90%")
```

@@

\noindent Figure \ref{fig:dotplot-8} helps to compare households through the heatmap (a) 
across harmony pairs with x-variable and facet variable plotted on the x-axis and y-axis of the heatmap. Here dom, dow, wdwnd are abbreviations for day-of-month, day-of-week and weekday/weekend and so on. (a) emphasizes patterns not discernible through (b), which is a linear display containing the raw data for one month (Sep-2019), with the major and minor x-axis corresponding to weeks and days respectively. In (a), the colors represent the value of $wpd$, implying darker cells correspond to more significant harmony pairs. Also, the ones with `*` corresponds to the ones above $wpd_{threshold95}$. This plot suggests that there are no significant periodic patterns for id 5. Household id 6 and 7 differ in the sense that for id 6, the difference in patterns only during weekday/weekends, whereas for id 7, all or few other days of the week are also important. This might be due to their flexible work routines or different day-off. id 7 and 8 have the same significant harmonies despite having very different total energy usage. Note that the $wpd$ values are computed over the entire range, but the linear display is zoomed into September.

@@

```{r dotplot-8, fig.cap = " A heatmap across harmonies (a) and line plot (b) showing energy usage for all household ids. The darker the colour in (a), the higher the harmony is ranked. Visualizing the harmony pairs in this way helps us to see the important cyclic granularities along the x-axis, facet or both. It further shows which households have significant patterns across similar temporal granularities. For eg, id 7 and 8 have significant patterns across (hod, dow) and (dow, hod), which was not apparent in (a).", fig.height = 8}
# elec_sig_split <- elec_harmony_all %>%
#   mutate(select_split =  str_split(select_harmony, " ", simplify = TRUE)[,2]) %>% 
#   
  select_split <- str_split(elec_harmony_all$select_harmony, " ", simplify = TRUE)[,2]

  elec_sig_split <- elec_harmony_all %>% 
    bind_cols(select_split = select_split) %>% 
  mutate(significant = case_when(
    select_split == "***" ~ "highest",
    select_split == "**" ~ "high",
    select_split == "*" ~ "medium",
    select_split == "" ~ "low"
  )) %>% 
   mutate(rank = case_when(
    select_split == "***" ~ paste(rank, "***", sep = " "),
    select_split == "**" ~  paste(rank, "**", sep = " "),
    select_split == "*" ~  paste(rank, "*", sep = " "),
    select_split == "" ~  paste(rank, "", sep = " ")
  ))

elec_sig_split$significant <- 
  factor(elec_sig_split$significant, levels = c("highest", "high", "medium", "low"))

heatplot <- elec_sig_split %>% 
  mutate(significance_95 = if_else(significant %in% c("high", "highest"), "*", "")) %>% 
  ggplot(aes(x = x_variable,
                 y = facet_variable)) +
  geom_tile(aes(fill = wpd)) + 
                #color = as.factor(significance_95)),
            #size = 0.3) +
  geom_text(aes(label = significance_95), color = "#42275a") +
  scale_fill_gradient(low = "white",high = "#ba5370") +
  #scale_fill_manual(palette = "Dark2") +
  #scale_colour_manual(values = c("white","red")) + 
  theme(legend.position = "bottom") +
  coord_fixed() + 
  guides(fill = guide_legend()) +
  theme_void() +
  theme_gray(base_size = 12, base_family = "Times") +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 60, hjust=1),legend.position = "bottom") + ggtitle("a") +
  theme(
    strip.background = element_blank(),
    strip.text.y  = element_blank(), plot.margin = unit(c(0, -2, 0, 0), "cm")) + ggtitle("(a)")

elec <- read_rds(here("paper/data/elec_all-8.rds")) %>% 
  dplyr::filter(date(reading_datetime) >= ymd("20190701"), date(reading_datetime) < ymd("20191231"), meter_id==1) %>% 
  select(-meter_id) %>% 
  rename("id" = "household_id",
         "date_time" = "reading_datetime")

# 
# elec %>% 
#   filter(id == "1") %>% 
#   prob_plot("hour_day", 
#             "day_week",
#             response = "kwh",
#             plot_type = "violin", fill = "blue")
#   
elec1 <- elec %>% 
  filter(id == "1") %>% 
  create_gran("day_week") %>% 
  create_gran("hour_day") %>% 
  mutate(day_week = as.numeric(day_week),
         hour_day = as.numeric(hour_day),
         id = as.numeric(id))

# temp.gly <- glyphs(elec1, "day_week", "hour_day", "id","kwh", height = 2.5)
# 
# ggplot(temp.gly, ggplot2::aes(gx, gy, group = gid)) +
#   add_ref_lines(temp.gly, color = "grey90") +
#   add_ref_boxes(temp.gly, color = "grey90") +
#   geom_path() +
#   theme_bw() +
#   labs(x = "", y = "")

elec_zoom <-  elec %>%
  as_tibble() %>% 
  dplyr::filter(date(date_time) > as.Date("2019-09-01") & date(date_time) < (as.Date("2019-09-30"))) %>%
  ggplot(aes(x=date_time, y = kwh)) +
  #geom_point(size = 0.1, colour = "black", alpha = 0.3) +
  geom_line(size = 0.1, colour = "blue") +
  facet_wrap(~id, 
             scales = "free_y",
             ncol = 1,
             strip.position =  "right",
             labeller = "label_both") + 
  xlab("Time [30m]") + 
  theme_grey() + 
  ylab("linear energy demand (kwh) for September") + ggtitle("(b)") +
  theme(panel.grid.major.x = element_blank()) +
  scale_x_datetime("Date", date_labels = "%b %d",
                   breaks = "1 week",
                   date_minor_breaks = "1 day")  + theme_bw() +
  theme(panel.grid.major.x =  element_line(colour = "#A9A9A9"),
        panel.grid.minor.x =  element_line(colour = "#D3D3D3"))

heatplot +  facet_grid(id~.) + elec_zoom +
  theme( plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  plot_layout(widths = c(1, 2)) 
```

```{r tab-rank-8}
elec_rank <- elec_sig_split %>% 
  select(-c(6,7, 9, 10)) %>% 
  pivot_wider(
              names_from = id,
                values_from = rank) %>% 
  rename("facet variable" = "facet_variable",
         "x variable" = "x_variable") %>% 
  select(-facet_levels, -x_levels)

elec_rank %>% kable(format = "markdown", caption = "Ranking of harmonies for the eight households with significance levels.")
```

@@

id4_tsibble <- elec %>%
  filter(id == 4)

id5_tsibble <- elec %>%
  filter(id == 5)

p1 <- id4_tsibble %>% 
   prob_plot("hour_day",
             "day_week",
             response = "kwh",
             plot_type = "quantile",
             symmetric = TRUE,
             quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)) +
  ggtitle("a) hod vs dow (id:4)") + 
  #scale_colour_brewer(name = "", palette = "Set2") +
  theme(legend.position = "none",
        strip.text = element_text(size = 7, margin = margin()),)

p2 <- id5_tsibble %>%
   prob_plot("hour_day",
             "day_week",
             response = "kwh",
             plot_type = "quantile",
             symmetric = TRUE,
             quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)) +
  ggtitle("a) hod vs dow (id:5)") +
  #scale_colour_brewer(name = "", palette = "Set2")   +
  theme(legend.position = "none",
        strip.text = element_text(size = 7, margin = margin()))

p1 + p2

>>>>

Exploratory data analysis, as coined by John W. Tukey [@Tukey1977-jx] involves many iterations of finding structures and patterns that allow the data to be informative. With temporal data available at finer scales, exploring time series data can become overwhelming with so many possible cyclic temporal granularities [@Bettini1998-ed, @Gupta2020-vo], which are temporal deconstructions that represent cyclic repetitions in time, e.g. hour-of-day, day-of-month, or any public holidays. These granularities form ordered (for eg. day-of-week where Tue is always followed by Wed, which again is followed by Thu and so on) or unordered categorical variables. Therefore, exploring univariate time series data amounts to exploring the distribution of the measured variable across different categories of the cyclic granularities. \noindent Take the example of the  electricity smart meter data used in @wang2020calendar for four households in Melbourne, Australia. Figure \ref{fig:onegran-new} shows the distribution of energy usage of one household (id 2) across cyclic granularity a) hour-of-day and b) month-of-year. Figure \ref{fig:onegran-new}(a) shows that energy consumption is higher during the morning hours (5-8) when members in the household wake up and then again in the evening hours (17-20) possibly when members get back from work with maximum variation (large inter-quartile range) in behavior in the afternoon hours (12-16). Figure \ref{fig:onegran-new}(b) shows the distribution of energy consumption across months January to June. The median and quartile deviation of energy usage in Jan and Feb are generally on a much higher side, possibly due to the usage of air conditioners (Jan, Feb are peak summer in Australia), however for other months (Mar-Jun, autumn and winter), the smaller median and quartile deviation indicate a more consistent behavior. It might also imply that this household do not use as much heater as compared to air conditioner. A lot of households in Victoria use gas heating and hence the usage of heaters might not be reflected here. Potentially many such displays could be drawn across day-of-week, day-of-month, weekday/weekend, or any other chosen cyclic granularities of interest. However, all of them would not be interesting to discern important patterns in the energy usage. Only those displays, which have "significant" distributional differences between categories of the cyclic granularity would be informative.

@@

```{r onegran-new, fig.cap="Boxplots showing the distribution of one household across one cyclic granularity at a time - (a) hour-of-day and (b) and month-of-year. The daily and annual periodic behavior of energy is apparent in (a) and (b) respectively, with daily peaks occuring in morning and evening hours when members in the house are present and active, more volatility (usage of air conditioner) in summer months (Jan, Feb) due to air conditioners and more consistent behavior in winter and autumn months (Mar, Apr, May, June)."}
```

<!-- comparing one and two grans for one households -->
\noindent Exploring the distribution of the measured variable across two cyclic granularities tend to provide more detailed information on its structure. For example, Figure \ref{fig:id2-new2}(a) slice down further by showing the usage distribution across hour-of-day conditional on month-of-year across two households (id 2 and 4). It shows the hourly usage over a day do not remain same across months. Unlike other months, the 75th and 90th percentile for all hours of the day in January are high, pretty close and are not characterized by a morning and evening peak. The household in Figure \ref{fig:id2-new2}(b) has 90th percentile consumption higher in summer months relative to autumn or winter, but the 75th and 90th percentile are far apart in all months, implying that the second household resorts to air conditioning much less regularly than the first one. The differences seem to be more prominent across month-of-year (facets) than hour-of-day (x-axis) for this household, whereas they are prominent for both cyclic granularities for the first household. It could be immensely useful to make the transition from all potential displays to the ones that are informative across atleast one cyclic granularity.

@@

<!-- \ref{fig:id2-new2} shows the distribution of energy consumption across hour-of-day conditional on month-of-year. \ref{fig:id2-new2}(a) shows that energy consumption is higher during the morning hours when members in the household wake up and then again higher in the evening hours possibly when members are back from work with them showing maximum variable behavior in the afternoon hours. \ref{fig:id2-new2}(b) shows the months Jan and Feb not only have a high median consumption, but also more variability, possibly due to the usage of air conditioning. The other months shows more consistent behavior with the energy consumption rising in May and June. \ref{fig:id2-new2} adds to this information, by showing that hours in Jan do not necessarily follow the same pattern across the day (due to Holidays, people visiting). Moreover, the volatility in consumption are almost equal and high for all hours in Jan, whereas variability is most in the evening hours for other months. The distribution of consumption across different hours of the day look similar for Apr, May and Jun, characterized by a distinct morning, afternoon and evening peaks and higher variability in evening hours. -->

@@

```{r id2-new2, fig.cap = "Distribution of energy consumption displayed through area quantile plots across two cyclic granularities month-of-year and hour-of-day and two households. The black line is the median, whereas the orange band covers the 25th to 75th percentile and the green band covers the 10th to 90th percentile. Difference between the 90th and 75th quantiles is less for (Jan, Feb) for the first household (a), suggesting that it is a more frequent user of air conditioner than the second household (b). Energy consumption for in (a) changes across both granularities, whereas for (b) daily pattern stays same irrespective of the months."}
```

@@

@@

\noindent Figure \ref{fig:dotplot-8} helps to compare households through the heatmap (a) 
across harmony pairs with the cyclic granularity mapped to x-axis and facet being plotted on the x-axis and y-axis of the heatmap. Here dom, dow, wdwnd are abbreviations for day-of-month, day-of-week and weekday/weekend and so on. (a) emphasizes patterns not discernible through (b), which is a linear display containing the raw data for one month (Sep-2019), with the major and minor x-axis corresponding to weeks and days respectively. In (a), the colors represent the value of $wpd$, implying darker cells correspond to more significant harmony pairs. Also, the ones with `*` corresponds to the ones above $wpd_{threshold95}$. This plot suggests that there are no significant periodic patterns for id 5. Household id 6 and 7 differ in the sense that for id 6, the difference in patterns only during weekday/weekends, whereas for id 7, all or few other days of the week are also important. This might be due to their flexible work routines or different day-off. id 7 and 8 have the same significant harmonies despite having very different total energy usage. Note that the $wpd$ values are computed over the entire range, but the linear display is zoomed into September.

@@

```{r dotplot-8, fig.cap = " A heatmap across harmonies (a) and line plot (b) showing energy usage for all household ids. The darker the colour in (a), the higher the harmony is ranked. Visualizing the harmony pairs in this way helps us to see the important cyclic granularities along the x-axis, facet or both. It further shows which households have significant patterns across similar temporal granularities. For eg, id 7 and 8 have significant patterns across (hod, dow) and (dow, hod), which was not apparent in (a).", fig.height = 8}
```

```{r tab-rank-8}
```

@@
